---
title: "Intersections Between Two Adjacent Categories in the Graded Response Model"
description: |
 The interpretation of step parameters in the Graded Response Model is different than the RSM/PCM/GPCM family due to a different functional form. While the step parameters in the RSM/PCM/GPCM family represent the point on the latent trait continuum where one category becomes more likely than the previous category, it is not same in the Graded Response Model. So, one curious student wondered at what point on the latent trait continuum the intersections occur between two adjacent categories for the Graded Response Model.
 
draft: true

author:
  - name: Cengiz Zopluoglu
    affiliation: University of Miami
    
date: 2-26-2020

categories:
  
  - "2020"
  - "item response theory"
  - "graded response model"
  
output:
  distill::distill_article:
    self_contained: true

preview: files/grm.jpeg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

One of the IRT models available for polytomous response outcomes (response outcomes with more than two categories) is the [Graded Response Model (GRM)](https://psycnet.apa.org/record/1972-04809-001). GRM can be used for questionnaires with common rating scale items which item responses are ordered categorical responses such as Strongly Disagree (0), Disagree (1), Agree (2), and Strongly Agree (3). 

In GRM, for an item with $K$ categories (k=0,1,2,...K), the probability of selecting a response category *j* (j=1,2,..K) or above is defined as:

$$
P(y\geq j)=\frac{1}{1+e^{a(\delta_j-\theta)}},
$$
where $a$ is the item discrimination, $\delta_j$ is the between-category threshold, and $\theta$ is the latent trait for the respondent. The number of the between-category thresholds for each item to be estimated is equal to K-1. The probability of selecting each response category is subsequently defined as:

$$
P(y=0) = 1 - P(y\geq 1)
$$
$$
P(y=1) = P(y\geq 1)  - P(y\geq 2)
$$
$$
P(y=2) = P(y\geq 2)  - P(y\geq 3)
$$
$$
...
$$
$$
P(y=K)=P(y\geq K).
$$

Below is a plot called Category Response Curves for an hypothetical 5-point rating scale item with parameters $a=2$, $\delta_1=-3$, $\delta_2=-2$, $\delta_3=0$, $\delta_4=1$. 

```{r eval=TRUE, echo=FALSE, fig.align='center',fig.height=6,fig.width=8,fig.cap='Graded Response Model Category Response Curves'}
a  = 2
d1 = -3
d2 = -2
d3 = 0
d4 = 1

theta <- seq(from=-4,to=4,.01)

p1orabove = 1/(1+exp(a*(d1-theta)))
p2orabove = 1/(1+exp(a*(d2-theta)))
p3orabove = 1/(1+exp(a*(d3-theta)))
p4orabove = 1/(1+exp(a*(d4-theta)))

p0 = 1-p1orabove
p1 = p1orabove - p2orabove
p2 = p2orabove - p3orabove
p3 = p3orabove - p4orabove
p4 = p4orabove

plot(theta,p0,type="l",ylim=c(0,1),xlab=expression(theta),ylab="P(y=k)",xaxt='n')
axis(side=1,at=seq(-4,4,1))
points(theta,p1,type="l")
points(theta,p2,type="l")
points(theta,p3,type="l")
points(theta,p4,type="l")
text(-4,.93,"Cat 0")
text(-2.5,.49,"Cat 1")
text(-1,.79,"Cat 2")
text(.5,.49,"Cat 3")
text(3.5,.95,"Cat 4")

#int1 = log(exp(a*d1+a*d2)/(exp(a*d2)-2*exp(a*d1)))/a
#int2 = log((exp(a*d1+a*d2)+exp(a*d2+a*d3)-(2*exp(a*d1+a*d3)))/(exp(a*d1)+exp(a*d3)-(2*exp(a*d2))))/a
#int3 = log((exp(a*d2+a*d3)+exp(a*d3+a*d4)-(2*exp(a*d2+a*d4)))/(exp(a*d2)+exp(a*d4)-(2*exp(a*d3))))/a
#int4 = log(exp(a*d4)-2*exp(a*d3))/a

#p1orabove = 1/(1+exp(a*(d1-int1)))
#p0 = 1-p1orabove
#points(int1,p0,type="p",pch=16,cex=1)

#p1orabove = 1/(1+exp(a*(d1-int2)))
#p2orabove = 1/(1+exp(a*(d2-int2)))
#p1 = p1orabove-p2orabove
#points(int2,p1,type="p",pch=16,cex=1)

#p2orabove = 1/(1+exp(a*(d2-int3)))
#p3orabove = 1/(1+exp(a*(d3-int3)))
#p2 = p2orabove-p3orabove
#points(int3,p2,type="p",pch=16,cex=1)

#p3orabove = 1/(1+exp(a*(d3-int4)))
#p4orabove = 1/(1+exp(a*(d4-int4)))
#p3 = p3orabove-p4orabove
#points(int4,p3,type="p",pch=16,cex=1)

abline(v=d1,lty=2,col="gray")
abline(v=d2,lty=2,col="gray")
abline(v=d3,lty=2,col="gray")
abline(v=d4,lty=2,col="gray")

p1orabove = 1/(1+exp(a*(d1-d2)))
p2orabove = 1/(1+exp(a*(d2-d2)))
p3orabove = 1/(1+exp(a*(d3-d2)))
p4orabove = 1/(1+exp(a*(d4-d2)))

p0 = 1-p1orabove
p1 = p1orabove-p2orabove
p2 = p2orabove-p3orabove
p3 = p3orabove-p4orabove
p4 = p4orabove

points(d2,p2,type="p",pch=16,cex=.8,col="blue")
points(d2,p3,type="p",pch=16,cex=.8,col="blue")
points(d2,p4,type="p",pch=16,cex=.8,col="blue")


p1orabove = 1/(1+exp(a*(d1-d3)))
p2orabove = 1/(1+exp(a*(d2-d3)))
p3orabove = 1/(1+exp(a*(d3-d3)))
p4orabove = 1/(1+exp(a*(d4-d3)))

p0 = 1-p1orabove
p1 = p1orabove-p2orabove
p2 = p2orabove-p3orabove
p3 = p3orabove-p4orabove
p4 = p4orabove

points(d3,p3,type="p",pch=16,cex=.8,col="blue")
points(d3,p4,type="p",pch=16,cex=.8,col="blue")

```


Category Response Curves display the probability of selecting each response category as a funciton of $\theta$. Vertical lines in Figure 1 represents the points on the latent trait continuum where $\theta=\delta_1$, $\theta=\delta_2$, $\theta=\delta_3$, and $\theta=\delta_4$. The interpretation of between-category threholds in GRM is that they represent the point on the latent trait continuum where the probability of responding to a certain category or above is equal to 0.5. For instance, $\delta_2=-2$  means that the probability of choosing Category 2 or above is equal to .5 when $\theta=-2$. In other words, the three blue dots on the vertical line where $\theta=\delta_2=-2$ represents the probability of responding Category 2, Category 3, and Category 4 and the sum of these three probabilities is equal to 0.5. Or, similarly, the two blue dots on the vertical line where $\theta=\delta_3=0$ represents the probability of choosing response category 3 and Category 4 and the sum is equal to 0.5. 

The interpretation of the category thresholds (also called as *step difficulty*) is different for a different family of models (Rating Scale Model, Partial Credit Model, and Generalized Partial Credit Model). In GPCM, we can write the probability of responding to a certain category in a direct way (and I think this is one of the ugliest equations in IRT and difficult to grasp):

$$
P(y=j)=\frac{e^{\sum_{h=0}^{j}a(\theta-\delta_j)}}{\sum_{j=0}^{K}e^{\sum_{h=0}^{j}a(\theta-\delta_j)}}
$$

For instance suppose we plot the Category Response Curves for the Generalized Partial Credit Model (GPCM) for the same set of parameters, $a=2$, $\delta_1=-3$, $\delta_2=-2$, $\delta_3=0$, $\delta_4=1$.  

```{r eval=TRUE, echo=FALSE, fig.align='center',fig.height=6,fig.width=8,fig.cap='Generalized Partial Credit Model Category Response Curves'}
a  = 2
d1 = -3
d2 = -2
d3 = 0
d4 = 1

theta <- seq(from=-4,to=4,.01)

exp0 = 1
exp1 = exp(a*(theta-d1))
exp2 = exp(a*(theta-d1)+a*(theta-d2))
exp3 = exp(a*(theta-d1)+a*(theta-d2)+a*(theta-d3))
exp4 = exp(a*(theta-d1)+a*(theta-d2)+a*(theta-d3)+a*(theta-d4))

p0 = exp0/(exp0+exp1+exp2+exp3+exp4)
p1 = exp1/(exp0+exp1+exp2+exp3+exp4)
p2 = exp2/(exp0+exp1+exp2+exp3+exp4)
p3 = exp3/(exp0+exp1+exp2+exp3+exp4)
p4 = exp4/(exp0+exp1+exp2+exp3+exp4)

plot(theta,p0,type="l",ylim=c(0,1),xlab=expression(theta),ylab="P(y=k)",xaxt='n')
axis(side=1,at=seq(-4,4,1))
points(theta,p1,type="l")
points(theta,p2,type="l")
points(theta,p3,type="l")
points(theta,p4,type="l")
text(-4,.93,"Cat 0")
text(-2.5,.6,"Cat 1")
text(-1,.81,"Cat 2")
text(.5,.61,"Cat 3")
text(3.5,.95,"Cat 4")
abline(v=d1,lty=2,col="gray")
abline(v=d2,lty=2,col="gray")
abline(v=d3,lty=2,col="gray")
abline(v=d4,lty=2,col="gray")
```




























<center>
<details>
  <summary>
  <font color="blue"><b> Click to see the R code for Figure 1 </b></font>
  </summary>
```{r eval=FALSE,echo=TRUE}
a  = 2
d1 = -2
d2 = -1
d3 = 0
d4 = 1

theta <- seq(from=-3,to=3,.01)

p1orabove = 1/(1+exp(a*(d1-theta)))
p2orabove = 1/(1+exp(a*(d2-theta)))
p3orabove = 1/(1+exp(a*(d3-theta)))
p4orabove = 1/(1+exp(a*(d4-theta)))

p0 = 1-p1orabove
p1 = p1orabove - p2orabove
p2 = p2orabove - p3orabove
p3 = p3orabove - p4orabove
p4 = p4orabove

plot(theta,p0,type="l",ylim=c(0,1),xlab=expression(theta),ylab="P(y=k)")
points(theta,p1,type="l")
points(theta,p2,type="l")
points(theta,p3,type="l")
points(theta,p4,type="l")
text(-3,.93,"Cat 0")
text(-1.5,.5,"Cat 1")
text(-.5,.5,"Cat 2")
text(.5,.5,"Cat 3")
text(2.6,.93,"Cat 4")
```
</details>
</center>
























